How has the Brazilian Amazon been constructed as a problem? The economy, the environment, the people, and the nation in the presidential speeches since 1985

### 2022-06-09 Draft_June_2022/Full_draft_20220608.tex

---
title: How has the Brazilian Amazon been constructed as a problem? The economy, the environment, the people, and the nation in the presidential speeches since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both co-authors contributed equally to the article; names are ordered alphabetically. Livio is a PhD Candidate in Anthropology and Sociology at the Geneva Graduate Institute and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences at the Geneva Graduate Institute and affiliated to the Centre for International Environmental Studies.]
date: "8 June 2022"
output:
  html_document:
    df_print: paged
  fig_caption: yes
  pdf_document: default
  word_document:
    reference_docx: template.docx
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: The Amazon Rainforest is one of the most important ecosystems in the world, and most of this biome is located in Brazil. Albeit the importance of understanding the relationship between the social construction of the Amazon and its effects on environmental outcomes, we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between, or within, governments. Discursive problem-construction within democracies legitimize ways of thinking and acting towards the environment while affecting decisions to deforest, or not, at the local level. Building on @hirschman1963 and @bacchi2009 conceptualization of chosen problems, we propose a framework to investigate how the Brazilian Amazon has been constructed as a problem in political discourse. Combining hand-coding and supervised machine learning, we classify statements about the Amazon forest or region in 6130 official presidential speeches since 1985. We find that, first, international environmental events appear to drive presidents to speak more about the Amazon. Second, while the Amazon as a problem of economic integration dominated discourses from 1985 to the mid-2000s, environmental conservation and social development constructions briefly surpasses economic integration from 2010 to 2015. Constructing the Amazon as an issue of sovereignty also became more pertinent from 2010 onwards. Finally, using a multinomial regression model, we find that presidents are more likely to construct the Amazon as a problem of environmental conservation, than economic integration or social development, as presidents move away from the Amazon region.
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(include = FALSE)
library(dplyr)
library(ggplot2)
library(tidyr)
library(skimr)
library(scales)
library(RColorBrewer)
library(nnet)
library(stargazer)
library(kableExtra)
library(tm)
library(tidytext)
library(topicmodels)
library(tibble)
Sys.setlocale("LC_ALL", "C")
options(stringsAsFactors = FALSE, scipen = 999)
```

```{r data, message=FALSE, warning=FALSE, include=FALSE}
# Load data
##### We are missing many observations for Collor from 1991 and 1992...
##### These are not in the repository for some reason...
##### Should we try and complete or should we at least disclose these in a footnote?
pop <- readRDS("~/Documents/GitHub/amazondef/Replication_files/BR_presid_speeches_final.Rds")
pop <- pop %>% select(-c(title, date, location))
ama <- readRDS("~/Documents/GitHub/amazondef/Replication_files/final_data_as.Rds")
ama <- ama %>% select(-c(title, date, text, false_positives)) %>% rename (ei= "EI", sd= "SD")
AAI <- readRDS("~/Documents/GitHub/amazondef/Replication_files//AAI.Rds")
amazon_def_year <- readRDS("~/Documents/GitHub/amazondef/Replication_files//amazon_def_year.Rds")
location_km <- read_excel("~/Documents/GitHub/amazondef/Replication_Files/location_km.xlsx") %>% 
  select(-location_cat)
ama <- left_join(ama, location_km, by = "location")
adr <- data.frame(year = c(1985, 1986, 1987), def_year = c(21.050, 21.050, 21.050)) # Missing year data
amazon_def_year <- rbind(adr, amazon_def_year) 
hp = data.frame(beg = c(1985, 1990, 1993, 1994, 2002, 2010, 2016, 2018),
                end = c(1990, 1993, 1994, 2002, 2010, 2016, 2018, 2022),
                label = c(1987, 1991, 1993, 1998, 2006, 2013, 2017, 2020),
                presid = c("Sarney", "Collor", "Franco", "Cardoso", "Lula",
                           "Rousseff", "Temer", "Bolsonaro"))
# Prepare data for problem constructions
sov <- ama %>% 
  filter(sov==1) %>% 
  count(sov, year) %>%
  group_by(year) %>%
  select(-c(sov))%>%
  rename (sov = n)
ei <- ama %>% 
  filter(ei==1) %>% 
  count(ei, year) %>%
  group_by(year) %>%
  select(-c(ei))%>%
  rename (ei = n) %>%
  ungroup() %>%
  tibble::add_row(year = 2017, ei = 0) %>%
  tibble::add_row(year = 2018, ei = 0)
sd <- ama %>% 
  filter(sd==1) %>% 
  count(sd, year) %>%
  group_by(year) %>%
  select(-c(sd))%>%
  rename (sd = n)
con <- ama %>% 
  filter(con==1) %>% 
  count(con, year) %>%
  group_by(year) %>%
  select(-c(con))%>%
  rename (con = n)
other <- ama %>% 
  filter(other==1) %>% 
  count(other, year) %>%
  group_by(year) %>%
  select(-c(other))%>%
  rename (other = n)
temp <- ei %>% 
  left_join(other) %>% 
  left_join(con) %>% 
  left_join(sd) %>% 
  left_join(sov) %>% 
  replace(is.na(.), 0)
# Get mixed types in
ama_mx <-  ama %>%
  unite(mixed_type, sov:con, sep="_") %>%
  mutate(mx_cat = factor(case_when(mixed_type == '1_0_0_0' ~ "Pure Sovereignty",
                                   mixed_type == '0_1_0_0' ~ "Pure Economic Integration", 
                                   mixed_type == '0_0_1_0' ~ "Pure Social Development",
                                   mixed_type == '0_0_0_1' ~ "Pure Conservation",
                                   mixed_type == '0_0_0_0' ~ "Other",
                                   mixed_type == '0_1_0_1' ~ "Economic Conservation",
                                   mixed_type == '0_1_1_0' ~ "Social Economic",
                                   mixed_type == '1_0_0_1' ~ "Sovereign Conservation",
                                   mixed_type == '0_1_1_1' ~ "Economic, Social,\nand Conservation",
                                   mixed_type == '1_1_0_0' ~ "Economic Sovereignty",
                                   mixed_type == '0_0_1_1' ~ "Social Conservation",
                                   grepl("1_0_1_0|1_0_1_0|1_0_1_1|1_1_0_1|1_1_1_0|1_1_1_1",
                                         mixed_type) ~ "Other Mixed-types"),
         levels = c("Pure Economic Integration", "Pure Conservation", "Pure Social Development",
                    "Pure Sovereignty", "Other", "Economic Conservation", 
                    "Social Economic", "Sovereign Conservation", "Economic, Social,\nand Conservation",
                    "Economic Sovereignty", "Social Conservation", "Other Mixed-types")),
         mx_cat2 = case_when(grepl("Pure Economic Integration|Pure Conservation|
                                   |Pure Social Development|Pure Sovereignty", mx_cat) ~ "Pure-types",
                             grepl("^Other$", mx_cat) ~ "Other",
                             grepl("Economic Conservation|Social Economic|Sovereign Conservation|
                                   |Economic, Social,\nand Conservation|Economic Sovereignty|
                                   |Social Conservation|Other Mixed-types",
                                   mx_cat) ~ "Mixed-types"),
         election_year = ifelse(grepl("1989|1994|1998|2002|2006|2010|2014|2018", year), 1, 0))
```

@@

# 1 Introduction

@@

Different governments along Brazilian history have been described as proponents of a specific view of the Amazon [@drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. While the military dictatorship is associated with both national sovereignty and then economic integration, Sarney's and Lula's presidencies are often tied to environmental conservation. However, governments are usually represented monolithically, advancing a view that specific governments see the Amazon as an instance of one specific problem. Albeit the current calls to understand the environment as a social-cultural construction and to identify the effect of culture on environmental outcomes [@lepolaindewaroux2021], we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between, or within, governments.

In this article, we investigate how the Brazilian Amazon has been constructed as a problem in political discourses. Building on @hirschman1963 concept of chosen problems and @bacchi2009 concept of problem-representation in policy, we propose a framework to identify problem-constructions in political discourses. Although problem-construction takes place in a series of instances, we analyze the case of discourses by Brazilian presidents since 1985. We opt for presidential speeches for two reasons. First, presidential discourses have the power to introduce and justify public policy, as well as shape its perception to broad audiences [@zarefsky2004; @gillion2016]. It legitimizes ways of thinking about the Amazon. In turn, policy perception is key for policy adoption and implementation [@alesina2009; @lopez2020]. This is especially pertinent to deforestation in Brazil as expectations from governments' response, generated from material and discursive governmental practices [@assuncao2015; @capobianco2019; @capobianco2021], are a crucial factor in decisions to deforest at the ground [@campbell2015]. Second, we argue that presidents can employ specific problem-constructions that build objects as specific problems depending on who they speak to. Presidential discourses take place in different sites, from launching a new bridge in a small municipality in the middle of the Amazon to a keynote speech in a business association in São Paulo or at the UN general assembly in New York. Working with presidential discourses allows us to identify this variation in meaningful ways and better understand how the Amazon is socially constructed.

@@

# 2 Conceptual framework and contribution

## 2.1 Chosen problems, presidential discourse, and policy objects

Governments often have a choice in how they construct problems and what policy solutions they implement. In "Journey towards Progress", @hirschman1963 draws a conceptual distinction between pressing problems (pressured from outside parties to the government) and chosen problems (chosen by the government at their own discretion). Pressing problems can be either privileged or neglected, depending on the degree of pressure exercised by the interested groups. Chosen problems are those governments select at their discretion. Problems can change from pressing to chosen across time and in space as a function of (a) solutions becoming available, (b) a change of level of government control in society, or (c) a shift of interests from top policymakers [@hirschman1963].

Additionally, @bacchi2009 [p. 10] argues that policies have a cultural dimension as "it takes shape within specific historical and national or international contexts". The existence or proposal of a policy implies that there is a (public) problem that needs (governmental) action to be fixed. The alleged problem is not always explicitly stated in policy: policies are represented as solutions for implicit problems.\@hirschman1975[p. 388] exemplifies chosen problems with the case of the construction of Brasilia. Building Brasilia could solve a problem of regional inequality, a problem of a dormant economy without state investment, a problem of political representation, or all three. It is up to the discretion of the government to highlight or not the implicit problem a policy solves. A same policy object (e.g. Brasilia, the Amazon, criminality, poverty), thus, can be constructed as pressing or not depending on how a government represents it.

Governments can choose to emphasize (or not) one or more implicit problems that a policy solves depending on which interest group they are in communication with. Relative to how the policy is represented to be, it can be a solution of problems that are considered pressing or not for different groups and it is up to the discretion of the political actor to construct a particular problem in a particular way given context. @putnam1988 seminal article on the two-level game can help us make sense of this variation conceptually. The author argues that the outcomes of international negotiations lie within the overlap between the pressure of domestic groups and international groups. Relatedly, @hochstetler2012 argue that international commitments in the Brazilian case shift as a function of domestic cost-benefit calculations, as the country as a large emitter does not abide to the global commons framings. Taking the example of the 2012 Forest Code and the Amazon, President Dilma could emphasize its rural character for environmentalists at the Paris Summit, or the amnesty for deforestation for agricultural elites. The same policy, then, can solve the problem of environmental conservation as well as economic integration.

This variation in problem-construction suggests that governments are more diverse in their positions as the literature suggests. We conceptualize the Amazon region, forest, and people to be a policy object, this is a specific issue that deserves dedicated policy. Several scholars describe federal governments' as proponents of specific policies for the Amazon [see @drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. In this literature, the 1964 military dictatorship is associated with securing national sovereignty in the region by populating it and integrating it to the national economy [@drummond2006]. The governments from the late 1980s up to 2009, are more associated with environmental conservation. The presidencies of Dilma and Temer are connected to the decline of environmental policies, while Bolsonaro to the complete dismantling of environmental policies. While these works are important to understand how different governments acted towards the Amazon, the classification of specific governments into periods or policy cycles represents them as monolithic: they associate specific governments to one specific view of the Amazon. The presidency of Lula, often associated with environmental conservation because of the unprecedented decrease in deforestation. At the same time, during his terms governmental credits to agricultural expansion and the cattle industry were 20 times more than the budget of the Ministry of the Environment for this entire period [@capobianco2021, p. 106]. That is, there is more variation within governments than the literature suggests.

An interesting site to study the relationship between varied problem-construction and policy issues is presidential speeches as they legitimize ways of thinking about an issue. While the relationship between presidential speeches and policy is not causal, they have the power to introduce and justify public policy, as well as shape its perception to broad audiences [@zarefsky2004; @gillion2016]. Even if presidential discourse might not persuade public opinion regarding a proposed solution for a problem [@eshbaugh2010; @rottinghaus2009; @krebs2007], they direct the range of what is included in national debates [@druckman2004]. As presidents speak in varied locations to varied audiences, it is very likely that they chose to frame similar issues differently.

The possibility of different problem-constructions at the level of presidential speeches has important consequences for democracy. If presidents can promote different problem-constructions in different places, it is important to understand to what extent the implemented agenda correlates with their local, federal, or international problem-constructions. @pacheco2019 proposes that the Amazon is key to understand Brazilian state and democracy. Specifically, the author states that the natural richness of the region has been instrumentally transformed in political support through resource exploration by different governments over the last centuries. The costs for said economic benefits are the livelihoods of indigenous and traditional populations and the ecosystems they reside in. Political stability, thus, can be seen as a product of the trade-off between both. Policies during the military dictatorship were strongly geared toward integrating the Amazon into the national territory and international economy. These were times when deforestation increased, but political stability was achieved as those bearing the costs of said deforestation were being targeted. Generally high levels of violence in the region, epitomized in the assassination of Chico Mendes, serve as evidence for that [@hecht1990]. With the return of democracy in the 1980s, violence could not be used at the same extent as before and policies had to respond to growing environmental concern.

Relatedly, the possibility of different problem-constructions has consequences for environmental outcomes as well. When Brazilian presidents speak about the Amazon it not only makes headlines, nationally and internationally [@brice2021; @harris2021; @miranda2021], but also incites responses, shapes expectations, and feeds into the behavior of many actors involved in the Amazon. Deforestation rates in Brazil are more responsive to the government's environmental policy, generated from material and discursive governmental practices, than exogenous factors as market fluctuations [@assuncao2015; @capobianco2019; @capobianco2021]. Specifically, Capobianco finds that that the perception of higher risk of being caught for deforestation in the Amazon is the main mechanism through which environmental policy is effective. Relatedly, ethnographic research has also shown that expectations about governments' response are a crucial factor in decisions to deforest at the ground [@campbell2015]. Thus, understanding diversity in presidential speeches is a first step to connect it to concrete environmental outcomes.

The connection between presidential discourse and the environment has been studied in the case of the United States but remains conceptually underdeveloped. @calderwood2019, for instance, examines 2919 mentions of climate change in American official presidential speeches since 1989. Among various findings, one that stands out is that American presidents frequently side-step the environmental aspects of climate change. He also identifies a shift from economic to security framing of climate issues, side-lining its environmental aspects. @calderwood2020 also tests the effect of geographic location and type of communication regarding climate change. Building prominently on @putnam1988, he hypothesizes that presidents are more likely to mention climate change in foreign locations, and that location influences the specific discursive approach and tone they adopt. The author finds evidence in support of his hypothesis, suggesting American presidential discourse at the top on global warming changes based on location. Elsewhere, @bevitori2015 investigates how the 'environment' has been constructed in American presidential discourse since 1960 using a more automated approach. The author finds that mentions of the environment are typically co-selected with the pronoun 'our', as well as with 'economy', 'clean', and 'preserve'. These studies corroborate the possibilities of variation depending on the audience, but we are yet to tie vocabulary to specific problem-constructions based on wider shared meanings of environment and climate change. While these findings are US specific, they indicate that presidents raise different points about the environment at local, national, or international settings, depending on who they assume their audience is at that specific instance. Nevertheless, as @bacchi2009, we argue that problem-constructions touch on shared meanings about the region that are available to the speaker as part of larger social-cultural history. Mentions to the lexicon as 'environment', 'preserve', or 'climate change' should be tied to the larger meaning of policies in the United States and elsewhere.

On the other hand, there is also an empirical gap in terms of how the Amazon has been constructed as a problem along time and geographic location. While governmental discourses in Brazil have been studied for topics such as inflation or race relations, we only find one systematic analysis of Amazonian discourse. @barros2020 investigates Amazonian discourse in the Brazilian Congress with the objective of identifying the arguments put forth by congressmen. The main finding is that the economic value of the Amazon for the cattle industry is the most salient narrative, leading the author to conclude there is a mismatch between the international debate (which focuses on preservation) and the national debate (which focuses on economic development). We also find one analysis of Bolsonaro's rhetoric about the Amazon, but the study relies on a reduced sample without transparent explanations of how speeches were selected [@marquardt2022].

We propose a conceptual framework that accepts the possibility of varied problem-construction for the same policy issue and connects it to presidential speeches. Governments choose what problems to solve and what policies to implement. The same policies can be represented as solving different problems implying a degree of social construction [@bacchi2009]. Governments can choose to emphasize or not one or more implicit problems that a policy solves depending on which interest group they are in communication with. The variation in problem-construction suggests that Brazilian governments are more diverse in their positions as the literature suggests. An interesting site to study the relationship between varied problem-construction and policy is presidential speeches as they legitimize ways of thinking about an issue, matter for democracy, and for environmental outcomes. As presidents speak in different places and to different audiences, it is an empirical site prone to identify variation.

## 2.2 Problem-construction in Amazonian historiography

@@

### National Sovereignty

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={How has the Brazilian Amazon been constructed as a problem? The economy, the environment, the people, and the nation in the presidential speeches since 1985},
  pdfauthor={Livio Silva-Muller; Henrique Sposito},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{floatrow}
\floatsetup[figure]{capposition=top}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{How has the Brazilian Amazon been constructed as a problem? The
economy, the environment, the people, and the nation in the presidential
speeches since 1985}
\author{Livio Silva-Muller \and Henrique Sposito\footnote{Both
  co-authors contributed equally to the article; names are ordered
  alphabetically. Livio is a PhD Candidate in Anthropology and Sociology
  at the Geneva Graduate Institute and affiliated to the Albert
  Hirschman Centre on Democracy. Henrique is a PhD Candidate in
  International Relations and Political Sciences at the Geneva Graduate
  Institute and affiliated to the Centre for International Environmental
  Studies.}}
\date{8 June 2022}

\begin{document}
\maketitle
\begin{abstract}
The Amazon Rainforest is one of the most important ecosystems in the
world, and most of this biome is located in Brazil. Albeit the
importance of understanding the relationship between the social
construction of the Amazon and its effects on environmental outcomes, we
lack empirical accounts of how the Brazilian Amazon has been constructed
as a problem in discourses over time, by geographical location, and
between, or within, governments. Discursive problem-construction within
democracies legitimize ways of thinking and acting towards the
environment while affecting decisions to deforest, or not, at the local
level. Building on Hirschman (1963) and Bacchi (2009) conceptualization
of chosen problems, we propose a framework to investigate how the
Brazilian Amazon has been constructed as a problem in political
discourse. Combining hand-coding and supervised machine learning, we
classify statements about the Amazon forest or region in 6130 official
presidential speeches since 1985. We find that, first, international
environmental events appear to drive presidents to speak more about the
Amazon. Second, while the Amazon as a problem of economic integration
dominated discourses from 1985 to the mid-2000s, environmental
conservation and social development constructions briefly surpasses
economic integration from 2010 to 2015. Constructing the Amazon as an
issue of sovereignty also became more pertinent from 2010 onwards.
Finally, using a multinomial regression model, we find that presidents
are more likely to construct the Amazon as a problem of environmental
conservation, than economic integration or social development, as
presidents move away from the Amazon region.
\end{abstract}

@@

\hypertarget{introduction}{%
\section{1 Introduction}\label{introduction}}

@@

Different governments along Brazilian history have been described as
proponents of a specific view of the Amazon (Drummond and Barros-Platiau
2006; Pádua 2012; Franchini and Viola 2019; Capobianco 2019; Pereira and
Viola 2021). While the military dictatorship is associated with both
national sovereignty and then economic integration, Sarney's and Lula's
presidencies are often tied to environmental conservation. However,
governments are usually represented monolithically, advancing a view
that specific governments see the Amazon as an instance of one specific
problem. Albeit the current calls to understand the environment as a
social-cultural construction and to identify the effect of culture on
environmental outcomes (Waroux et al. 2021), we lack empirical accounts
of how the Brazilian Amazon has been constructed as a problem in
discourses over time, by geographical location, and between, or within,
governments.

In this article, we investigate how the Brazilian Amazon has been
constructed as a problem in political discourses. Building on Hirschman
(1963) concept of chosen problems and Bacchi (2009) concept of
problem-representation in policy, we propose a framework to identify
problem-constructions in political discourses. Although
problem-construction takes place in a series of instances, we analyze
the case of discourses by Brazilian presidents since 1985. We opt for
presidential speeches for two reasons. First, presidential discourses
have the power to introduce and justify public policy, as well as shape
its perception to broad audiences (Zarefsky 2004; Gillion 2016). It
legitimizes ways of thinking about the Amazon. In turn, policy
perception is key for policy adoption and implementation (Alesina and
Giuliano 2009; López et al. 2020). This is especially pertinent to
deforestation in Brazil as expectations from governments' response,
generated from material and discursive governmental practices (Assunção,
Gandour, and Rocha 2015; Capobianco 2019, 2021), are a crucial factor in
decisions to deforest at the ground (Campbell 2015). Second, we argue
that presidents can employ specific problem-constructions that build
objects as specific problems depending on who they speak to.
Presidential discourses take place in different sites, from launching a
new bridge in a small municipality in the middle of the Amazon to a
keynote speech in a business association in São Paulo or at the UN
general assembly in New York. Working with presidential discourses
allows us to identify this variation in meaningful ways and better
understand how the Amazon is socially constructed.

@@

\hypertarget{conceptual-framework-and-contribution}{%
\section{2 Conceptual framework and
contribution}\label{conceptual-framework-and-contribution}}

\hypertarget{chosen-problems-presidential-discourse-and-policy-objects}{%
\subsection{2.1 Chosen problems, presidential discourse, and policy
objects}\label{chosen-problems-presidential-discourse-and-policy-objects}}

Governments often have a choice in how they construct problems and what
policy solutions they implement. In ``Journey towards Progress'',
Hirschman (1963) draws a conceptual distinction between pressing
problems (pressured from outside parties to the government) and chosen
problems (chosen by the government at their own discretion). Pressing
problems can be either privileged or neglected, depending on the degree
of pressure exercised by the interested groups. Chosen problems are
those governments select at their discretion. Problems can change from
pressing to chosen across time and in space as a function of (a)
solutions becoming available, (b) a change of level of government
control in society, or (c) a shift of interests from top policymakers
(Hirschman 1963).

Additionally, Bacchi (2009, 10) argues that policies have a cultural
dimension as ``it takes shape within specific historical and national or
international contexts''. The existence or proposal of a policy implies
that there is a (public) problem that needs (governmental) action to be
fixed. The alleged problem is not always explicitly stated in policy:
policies are represented as solutions for implicit
problems.@hirschman1975{[}p.~388{]} exemplifies chosen problems with the
case of the construction of Brasilia. Building Brasilia could solve a
problem of regional inequality, a problem of a dormant economy without
state investment, a problem of political representation, or all three.
It is up to the discretion of the government to highlight or not the
implicit problem a policy solves. A same policy object (e.g.~Brasilia,
the Amazon, criminality, poverty), thus, can be constructed as pressing
or not depending on how a government represents it.

Governments can choose to emphasize (or not) one or more implicit
problems that a policy solves depending on which interest group they are
in communication with. Relative to how the policy is represented to be,
it can be a solution of problems that are considered pressing or not for
different groups and it is up to the discretion of the political actor
to construct a particular problem in a particular way given context.
Putnam (1988) seminal article on the two-level game can help us make
sense of this variation conceptually. The author argues that the
outcomes of international negotiations lie within the overlap between
the pressure of domestic groups and international groups. Taking the
example of the 2012 Forest Code and the Amazon, President Dilma could
emphasize its rural character for environmentalists at the Paris Summit,
or the amnesty for deforestation for agricultural elites. The same
policy, then, can solve the problem of environmental conservation as
well as economic integration.

This variation in problem-construction suggests that governments are
more diverse in their positions as the literature suggests. We
conceptualize the Amazon region, forest, and people to be a policy
object, this is a specific issue that deserves dedicated policy. Several
scholars describe federal governments' as proponents of specific
policies for the Amazon (see Drummond and Barros-Platiau 2006; Pádua
2012; Franchini and Viola 2019; Capobianco 2019; Pereira and Viola
2021). In this literature, the 1964 military dictatorship is associated
with securing national sovereignty in the region by populating it and
integrating it to the national economy (Drummond and Barros-Platiau
2006). The governments from the late 1980s up to 2009, are more
associated with environmental conservation. The presidencies of Dilma
and Temer are connected to the decline of environmental policies, while
Bolsonaro to the complete dismantling of environmental policies. While
these works are important to understand how different governments acted
towards the Amazon, the classification of specific governments into
periods or policy cycles represents them as monolithic: they associate
specific governments to one specific view of the Amazon. The presidency
of Lula, often associated with environmental conservation because of the
unprecedented decrease in deforestation. At the same time, during his
terms governmental credits to agricultural expansion and the cattle
industry were 20 times more than the budget of the Ministry of the
Environment for this entire period (Capobianco 2021, 106). That is,
there is more variation within governments than the literature suggests.

An interesting site to study the relationship between varied
problem-construction and policy issues is presidential speeches as they
legitimize ways of thinking about an issue. While the relationship
between presidential speeches and policy is not causal, they have the
power to introduce and justify public policy, as well as shape its
perception to broad audiences (Zarefsky 2004; Gillion 2016). Even if
presidential discourse might not persuade public opinion regarding a
proposed solution for a problem (Eshbaugh-Soha 2010; Rottinghaus 2009;
Krebs and Jackson 2007), they direct the range of what is included in
national debates (Druckman and Holmes 2004). As presidents speak in
varied locations to varied audiences, it is very likely that they chose
to frame similar issues differently.

The possibility of different problem-constructions at the level of
presidential speeches has important consequences for democracy. If
presidents can promote different problem-constructions in different
places, it is important to understand to what extent the implemented
agenda correlates with their local, federal, or international
problem-constructions. Pacheco (2019) proposes that the Amazon is key to
understand Brazilian state and democracy. Specifically, the author
states that the natural richness of the region has been instrumentally
transformed in political support through resource exploration by
different governments over the last centuries. The costs for said
economic benefits are the livelihoods of indigenous and traditional
populations and the ecosystems they reside in. Political stability,
thus, can be seen as a product of the trade-off between both. Policies
during the military dictatorship were strongly geared toward integrating
the Amazon into the national territory and international economy. These
were times when deforestation increased, but political stability was
achieved as those bearing the costs of said deforestation were being
targeted. Generally high levels of violence in the region, epitomized in
the assassination of Chico Mendes, serve as evidence for that (Hecht and
Cockburn 1990). With the return of democracy in the 1980s, violence
could not be used at the same extent as before and policies had to
respond to growing environmental concern.

Relatedly, the possibility of different problem-constructions has
consequences for environmental outcomes as well. When Brazilian
presidents speak about the Amazon it not only makes headlines,
nationally and internationally (Brice and Smith 2021; Harris 2021;
Miranda 2021), but also incites responses, shapes expectations, and
feeds into the behavior of many actors involved in the Amazon.
Deforestation rates in Brazil are more responsive to the government's
environmental policy, generated from material and discursive
governmental practices, than exogenous factors as market fluctuations
(Assunção, Gandour, and Rocha 2015; Capobianco 2019, 2021).
Specifically, Capobianco finds that that the perception of higher risk
of being caught for deforestation in the Amazon is the main mechanism
through which environmental policy is effective. Relatedly, ethnographic
research has also shown that expectations about governments' response
are a crucial factor in decisions to deforest at the ground (Campbell
2015). Thus, understanding diversity in presidential speeches is a first
step to connect it to concrete environmental outcomes.

The connection between presidential discourse and the environment has
been studied in the case of the United States but remains conceptually
underdeveloped. Calderwood (2019), for instance, examines 2919 mentions
of climate change in American official presidential speeches since 1989.
Among various findings, one that stands out is that American presidents
frequently side-step the environmental aspects of climate change. He
also identifies a shift from economic to security framing of climate
issues, side-lining its environmental aspects. Calderwood (2020) also
tests the effect of geographic location and type of communication
regarding climate change. Building prominently on Putnam (1988), he
hypothesizes that presidents are more likely to mention climate change
in foreign locations, and that location influences the specific
discursive approach and tone they adopt. The author finds evidence in
support of his hypothesis, suggesting American presidential discourse at
the top on global warming changes based on location. Elsewhere, Bevitori
(2015) investigates how the `environment' has been constructed in
American presidential discourse since 1960 using a more automated
approach. The author finds that mentions of the environment are
typically co-selected with the pronoun `our', as well as with `economy',
`clean', and `preserve'. These studies corroborate the possibilities of
variation depending on the audience, but we are yet to tie vocabulary to
specific problem-constructions based on wider shared meanings of
environment and climate change. While these findings are US specific,
they indicate that presidents raise different points about the
environment at local, national, or international settings, depending on
who they assume their audience is at that specific instance.
Nevertheless, as Bacchi (2009), we argue that problem-constructions
touch on shared meanings about the region that are available to the
speaker as part of larger social-cultural history. Mentions to the
lexicon as `environment', `preserve', or `climate change' should be tied
to the larger meaning of policies in the United States and elsewhere.

On the other hand, there is also an empirical gap in terms of how the
Amazon has been constructed as a problem along time and geographic
location. While governmental discourses in Brazil have been studied for
topics such as inflation or race relations, we only find one systematic
analysis of Amazonian discourse. Barros (2020) investigates Amazonian
discourse in the Brazilian Congress with the objective of identifying
the arguments put forth by congressmen. The main finding is that the
economic value of the Amazon for the cattle industry is the most salient
narrative, leading the author to conclude there is a mismatch between
the international debate (which focuses on preservation) and the
national debate (which focuses on economic development).

We propose a conceptual framework that accepts the possibility of varied
problem-construction for the same policy issue and connects it to
presidential speeches. Governments choose what problems to solve and
what policies to implement. The same policies can be represented as
solving different problems implying a degree of social construction
(Bacchi 2009). Governments can choose to emphasize or not one or more
implicit problems that a policy solves depending on which interest group
they are in communication with. The variation in problem-construction
suggests that Brazilian governments are more diverse in their positions
as the literature suggests. An interesting site to study the
relationship between varied problem-construction and policy is
presidential speeches as they legitimize ways of thinking about an
issue, matter for democracy, and for environmental outcomes. As
presidents speak in different places and to different audiences, it is
an empirical site prone to identify variation.

\hypertarget{problem-construction-in-amazonian-historiography}{%
\subsection{2.2 Problem-construction in Amazonian
historiography}\label{problem-construction-in-amazonian-historiography}}

@@

\hypertarget{national-sovereignty}{%
\subsubsection{National Sovereignty}\label{national-sovereignty}}
